Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system with plausible mechanism

### 2024-07-31 index.qmd
section:background,dag; process:write,figures

@@

  - directed acycilc graph
  - causal analysis
  - probabilistic statistics

@@

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data could be generated. This study addresses these issues by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates this model into a probabilistic statistical framework to estimate statistical relationships and infer causal effects within the system. This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

@@

<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
Regardless of complexity, DAGs can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. Each panel of @fig-dags illustrates these building blocks. @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$. Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$. @fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$. Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$, and that the relationship between these variables is entirely mediated by $X_{2}$. @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.

@@

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, the use of fundamental building block to construct a causal structure of a system is easier to understand using a motivating example. The motivating example can also serve to signal about other conventions when using DAGs. Consider a system where variables $X$ and $Z$ influence a third variable $Y$. In this system, it is assumed that $X$ and $Z$ are dependent on their own processes and are, therefore, independent from each other. @fig-dag_example1 presents the plausible causal structure of this system. The DAG shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed. The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows. Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly. These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature. Although this DAG explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020].

<!-- emphasize that everything can be observed -->
<!-- change the motivating example to include all building blocks  -->

@@

![Full DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=100%}

![Simplified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=40%}

@@

@@

@@

@@

@@

@@

@@

## Additional definitions {#sec-appA}

## Why do we need to estimate judges' abilities? {#sec-appB}

@@

>>>>

  - causal inference
  - probability
  - Thurstone

@@

  - directed acyclic graph
  - structural causal models
  - statistical modeling

@@

  Dichotomous Comparative Judgment (DCJ) requires judges to compare pairs of stimuli to determine which one exhibits a higher degree of a specific trait. DCJ has proven effective and reliable across various fields [@Pollitt_2012b; @Jones_2015; @vanDaal_et_al_2016; @Bartholomew_et_al_2018; @Lesterhuis_2018; @Bartholomew_et_al_2020; @Marshall_et_al_2020; @Boonen_et_al_2020]. However, despite the method's widespread use, existing literature lacks a clear explanation of the complexities and assumptions underpinning the DCJ system, as well as the plausible mechanisms through which DCJ data could be generated. This study addresses these issues by representing DCJ within the framework of causal inference. Specifically, utilizing the structural approach, the study develops a scientific model to clarify plausible causal assumptions and mechanisms inherent in the DCJ system. The study then translates this model into a probabilistic statistical model to estimate statistical relationships and infer causal effects within the system. This research provides a robust probabilistic foundation for the statistical analysis of DCJ data, building upon Thurstone’s law of comparative judgment [-@Thurstone_1927]. Its findings offer valuable insights for researchers and analysts designing and implementing DCJ experiments.

@@

This section introduces fundamental concepts in causal inference, such as directed acyclic graphs (DAGs), structural causal models (SCMs), and the flow of association and causation in graphs. It is not a comprehensive description of causal inference methods. Readers interested in deeper exploration should consult introductory papers like @Pearl_2010, @Rohrer_2018, @Pearl_2019, and @Cinelli_et_al_2020. They may also find introductory books such as @Pearl_et_al_2018, @Neal_2020 and @McElreath_2020 useful. For more advanced study, seminal intermediate papers like @Neyman_et_al_1923, @Rubin_1974, @Spirtes_et_al_1991, and @Sekhon_2009, as well as books such as @Pearl_2009, @Morgan_et_al_2014 and @Hernan_et_al_2020 are recommended.

@@

<!-- 3. How do we draw DAGs? and what are the building blocks of DAGs? -->
One key advantage of DAGs is that, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. @fig-dags illustrates these building blocks. @fig-dag_bb1 depicts two unconnected nodes, representing an scenario where variables $X_{1}$ and $X_{2}$ are not causally related. @fig-dag_bb2 shows two connected nodes, illustrating a scenario where a parent node $X_{1}$ exerts a causal influence on a child node $X_{2}$. Consequently, $X_{2}$ is considered a *descendant* of $X_{1}$. @fig-dag_bb3 depicts a *chain* (or *pipe*), where $X_{1}$ influences $X_{2}$, and $X_{2}$ influences $X_{3}$. In this configuration, $X_{1}$ is a parent node of $X_{2}$, and $X_{2}$ a parent node of $X_{3}$. Furthermore, the DAG shows that $X_{1}$ is an *ancestor* of $X_{3}$ and that the relationship between these variables is entirely mediated by $X_{2}$. @fig-dag_bb4 illustrates a *fork*, where variables $X_{1}$ and $X_{3}$ are both influenced by $X_{2}$. In this scenario $X_{2}$ is a parent node of both $X_{1}$ and $X_{3}$. Finally, @fig-dag_bb5 depicts a *collider*, also known as *inmorality*, where variables $X_{1}$ and $X_{3}$ are concurrent causes of $X_{2}$. In this configuration, $X_{1}$ and $X_{3}$ are not causally related to each other but both influence $X_{2}$.

@@

<!-- 4. A motivating example -->
Given the heuristic nature of DAGs, a motivating example can illustrate how these five fundamental building blocks can be used to construct the causal structure of a system. This example can also help clarify additional rules and conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable $T$ on an outcome $Y$ needs to be investigated. Additionally, the problem indicates that a variable $X$ potentially influences both $Y$ and $T$. Such scenarios are not hard to imagine. For instance, $T$ could describe different treatments potentially influencing the binary onset of cancer ($Y$) while $X$ could denote the age of a patient. Using an example from DCJ, [we need an example]{style="color:red;"}. @fig-example presents two graphs describing the plausible causal structure of this system, the *simplified* and the *magnified* DAG.

<!-- 5. DAG for the motivating example -->
<!-- @fig-dag_example1 shows the endogenous variables $V=\{X,Z,Y\}$ as circled black nodes, indicating these variables are observed. The arrows connecting the variables indicate the direction of causal influence, while a lack of influence is indicated by the absence of arrows. Moreover, the exogenous variables $E=\{e_{X},e_{Z},e_{Y}\}$ represent everything else that is chosen not to be modeled explicitly. These exogenous variables, or *disturbances*, are usually represented by open circles to indicate their unobserved nature. Although this DAG, also known as Magnified DAG, explicitly shows the exogenous variables, conventionally these are omitted for brevity, resulting in an equivalent graph as shown in @fig-dag_example2 [@Cinelli_et_al_2020]. Sometimes it is convenient to show magnified graphs to show porblems in conditioning. -->

<!-- emphasize that everything can be observed -->
<!-- point out that the disturbances are usually stochastic in nature -->

@@

![Simplied DAG](images/figures/dag_example1.png){#fig-dag_example1 fig-align="center" width=45%}

![Magnified DAG](images/figures/dag_example2.png){#fig-dag_example2 fig-align="center" width=85%}

@@

<!-- (Structural Causal Model (SCM)) A structural causal model is a tuple of the following sets: -->
<!-- 1. A set ofendogenous variables 푉 2. A set ofexogenous variables 푈 3. A set of functions 푓, one to generate each endogenous variable as a function ofother variables Neal -->

@@

<!-- these are non-parametrical structural equations -->

<!-- Causal mechanism as a result of markov and minimality assumption -->
<!-- Modularity assumption also -->

@@

<!-- Mention convenient R package daggitty to draw dags and extract the causal assumption inherent in a graph -->

<!-- After detailing how to build graphs and what they imply, Include steps to draw dags using a motivating example -->

@@

<!-- The previous chapter gives graphical intuition for causal models, but it doesn’t explain how to identify causal quantities and formalize causal models. Neal -->

@@

<!-- Ardmed with dags and scm, and theit probabilité implication, we can now expand on thé motivating example provided in section xx -->

@@

<!-- no fundamental ontological objections to using DAGs to describe causal models and causal relationships. -->

@@

## Why do we need to estimate judges' abilities? {#sec-appA}

@@

## Latent variables as a mean of imputation {#sec-appB}

<!-- Latent variables and random effects are in essence imputation? (McElreath, p. 504) -->
<!-- Judge index works as a descendant of the latent variables, or the latent variable is not a reflective, but generative where the index is multiplied by a loading? -->
<!-- see Morgan et al (2014) -->
<!-- see Mayer (2019) -->

## Other comparative scenarios  {#sec-appC}
